Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-12-21 analysis/paper/shceof.tex
Actualiza

A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
Examples of this methodology are the SAM Index of \citet{gong1999}, the SH wave 3 activity index defined by \citet{raphael2004} and the SH zonally asymmetric circulation index from \citet{hobbs2010}.
These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
It is commonly easy the computation of these kind of indices, but they could be unable to capture non-stationary patterns.

@@

We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America \citep{cazes-boezio2003} and SH stratosphere-troposphere interactions \citep{lim2018}.

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and we computed horizontal wave activity fluxes following \citet{plumb1985}.
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2\degree and 2.5\degree horizontal resolution respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2022}.
Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures \citep{horel1984}.
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
The real and imaginary part of each spatial pattern represent the two phases of wave-like spatial patterns that are in quadrature.
The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & cEOF1 & cEOF2 & cEOF3\\
\midrule
cEOF1 & \textbf{0.29} & 0.01 & 0.03\\
cEOF2 & 0.00 & \textbf{0.59} & 0.02\\
cEOF3 & 0.00 & 0.00 & 0.01\\
\bottomrule
\end{tabular}
\end{table}

The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

@@

The phase of cEOFs is defined up to an additive constant.
For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part is minimised, which also nearly maximises the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables (e.g.~geopotential height, temperature, precipitation, and others).
For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
To obtain the linear coefficients of a variable \(X\) with the imaginary and real parts of each cEOF we fit the equation

\[
X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
\]

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for imaginary and real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to real (imaginary) part. Arbitrary units. The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.}
\end{figure}

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa. cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components. Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
\end{figure*}

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
The first mode (cEOF1) explains 82\% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7\%).
In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the standard EOFs of Figure~\ref{fig:eof-naive}.

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is more evident at 200 hPa.
There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

There is no significant simultaneous correlation between cEOFs time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean.
However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.

A significant positive trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, \citet{raphael2003} detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.

@@

In the previous section, cEOFs analysis was applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-gh-1.pdf}
\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa. These coefficients come from multiple linear regression involving the real and imaginary parts. Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
\end{figure}

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase \citep[e.g.][]{fogt2020}.
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
In the troposphere (Fig.~\ref{fig:eof2-regr-gh} column b), wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns \citep{mo2001}.
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} column a) supporting their equivalent barotropic nature.
But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2.
This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.}
\end{figure}

The signature of cEOFs variability on air temperature was also evaluated.
Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.

Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere \citep{hartmann1979, wirth1993, smith1995}.
In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.

>>>>

A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
Examples of this methodology are the SAM Index of \citet{gong1999}, the SH wave 3 activity index defined by \citet{raphael2004} and the SH zonally asymmetric circulation index from \citet{hobbs2010}.
These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
These kinds of indices are generally easy to compute, but they usually capture non-stationary patterns.

@@

We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2, the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and we computed horizontal wave activity fluxes following \citet{plumb1985}.
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2\degree and 2.5\degree horizontal resolution respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting monthly geopotential height fields onto the zonally asymmetric and zonally symmetric parts of the SAM spatial pattern, as proposed by \citet{campitelli2022}.
Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth EOFs of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).

To describe the phase-varying nature of these two waves one could combine each pair into indices of amplitude of phase.
So, for instance, the amplitude of the wave 1 EOF could be measured as \(\sqrt{\mathrm{EOF1}^2 + \mathrm{EOF2}^2}\) and it's phase as \(\tan^{-1} \left ( \frac{\mathrm{EOF2}}{\mathrm{EOF1}} \right )\) (where \(\mathrm{EOF1}\) and \(\mathrm{EOF2}\) are the time series associated with each EOF).
However, this rests upon visual inspection of the spatial patterns and only works if both phases appear cleanly in different EOFs, which is not guaranteed.
In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).

In Complex Empirical Orthogonal Functions (cEOF) analysis, each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series.
This is analogous to how any sine wave with any amplitude and phase can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase.
This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.

For example, Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of a cEOF, one with shaded contours (which we call the 0º phase) and the other with black contours (which we call the 90º phase).
This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern.
When the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.\\
Cases will have non zero values in both time series will represent intermediate phases.

These two phases are not special and the same cEOF could be represented by any two linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
This parallels traditional EOFs, which are defined up to an additive constant.
For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.

Compute cEOFs are computed the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
Each cEOF is a set of complex-valued spatial patterns and time series in which the real and imaginary parts are the two orthogonal phases of the wave-like pattern.

In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.

The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(r^2\)) between the time series of the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & cEOF1 & cEOF2 & cEOF3\\
\midrule
cEOF1 & \textbf{0.29} & 0.01 & 0.03\\
cEOF2 & 0.00 & \textbf{0.59} & 0.02\\
cEOF3 & 0.00 & 0.00 & 0.01\\
\bottomrule
\end{tabular}
\end{table}

The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

@@

As mentioned before, the choice of phases is arbitrary and equally valid.
But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.

In Section \ref{precipitation} we show regressions of precipitation and temperature associated with intermediate phases.
For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by \(\cos(\pi/4) + i\sin(\pi/4)\) and computing the regression on those rotated timeseries.

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between 75°S and 45°S.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables (e.g.~geopotential height, temperature, precipitation, and others).
For each cEOF, we computed regression maps by fitting a multiple linear model involving both the 0º and the 90º phases.
To obtain the linear coefficients of a variable \(X\) with the 0º and 90º phase of each cEOF we fit the equation

\[
X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0º}} + \beta(\lambda, \phi) \operatorname{cEOF_{90º}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
\]

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for 0º and 90º phases respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to 0º (90º) phase. Arbitrary units. The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.}
\end{figure}

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa. cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase. Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
\end{figure*}

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
The first mode (cEOF1) explains 82\% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7\%).
In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the 0º and the 90º phases are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the standard EOFs of Figure~\ref{fig:eof-naive}.

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (11\% vs.~3\%, respectively).
There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

There is no significant simultaneous correlation between cEOFs time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean.
However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
This is consistent with the fact that the mean zonally anomalous field of geopotential height is very similar to the cEOF1 (\(r^2\) = 98\%) and not similar to the cEOF2 (\(r^2\) = 0\%).

A significant positive trend in the 0º phase of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the phases of cEOF2.
The positive trend in the 0º cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.

@@

In the previous section, cEOFs analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-gh-1.pdf}
\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) 0º and (column 2) 90º phases of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa. These coefficients come from multiple linear regression involving the 0º and 90º phases. Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
\end{figure}

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the 0º cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with its centre shifted slightly eastward compared with the upper-level anomalies.
The 90º cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase \citep[e.g.][]{fogt2020}.
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
In the troposphere (Fig.~\ref{fig:eof2-regr-gh} row b) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}).
Regressed anomalies associated with the 0º cEOF2 are 1/4 wavelength out of phase with those associated with the 90º cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns \citep{mo2001}.
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.

These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} row a) supporting their equivalent barotropic nature.
But also present is a monopole over the pole with negative sign associated with the 0º cEOF2 and positive sign associated with the 90º cEOF2.
This monopole might indicate strengthening of the polar vortex associated with positive values of the 0º cEOF2 and weakening associated with negative values of 0º cEOF2.

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) 0º and (b) 90º phase of the cEOF1 for the 1979 -- 2019 period.}
\end{figure}

The signature of cEOFs variability on air temperature was also evaluated.
Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
In both levels, the 0º cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
On the other hand, the regression maps on the 90º cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.

Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both 0º and 90º phases throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere.
This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere \citep{hartmann1979, wirth1993, smith1995}.
In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 1/4 wavelength out of phase with horizontal and vertical transport, which are in addition 1/4 wavelength out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.
